---
title: "Experimental design for sequencing-based spatial transcriptomics with spaDesign"
author: "Juan Xie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental design for sequencing-based spatial transcriptomics with spaDesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rgl.useNULL = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Overview

`spaDesign` is a simulation-based framework for experimental design in
sequencing-based spatial transcriptomics studies. The package is designed to
help answer a practical question before committing to a full experiment:

> How much sequencing depth is enough to recover the spatial domains of interest?

Starting from a pilot spatial transcriptomics dataset, `spaDesign` learns
domain-specific expression and spatial patterns, simulates new datasets under
candidate sequencing depths and signal strengths, evaluates domain recovery, and
detects the sequencing depth at which additional reads provide diminishing
returns.

The typical workflow is:

1. Create a `spaDesign` object from a count matrix and spatial metadata.
2. Select domain-informative genes.
3. Estimate the spatial distribution and spatial expression models.
4. Simulate data across sequencing depth and perturbation settings.
5. Evaluate domain recovery with normalized mutual information (NMI).
6. Detect and visualize the saturation sequencing depth.

## Installation and setup

Install the development version from GitHub:

```{r install}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("JuanXie19/spaDesign")
```

Load the package:

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
suppressWarnings(suppressPackageStartupMessages(library(spaDesign)))
```

## Input data

The core input is a gene-by-spot count matrix and a spot-level metadata table.
The metadata must contain at least three columns:

- `x`: spatial x coordinate
- `y`: spatial y coordinate
- `domain`: known or estimated spatial domain label

The column names of the count matrix must match the row names of the metadata.

```{r input-format}
dim(count_matrix)
head(loc)

all(colnames(count_matrix) == rownames(loc))
```

`spaDesign` also ships example/reference data under `inst/extdata`. A small
example object can be loaded as follows:

```{r example-data, eval=TRUE}
example_path <- system.file("extdata", "exampleData.rds", package = "spaDesign")
example_data <- readRDS(example_path)

count_matrix <- example_data$count
loc <- example_data$loc

dim(count_matrix)
head(loc)
```

## Create a spaDesign object

Use `createDesignObject()` to initialize the S4 object used throughout the
analysis.

```{r create-object, eval=TRUE}
DATA <- createDesignObject(
  count_matrix = count_matrix,
  loc = loc
)

DATA
```

The object stores the reference count matrix, spot metadata, simulated data,
model parameters, selected genes, and downstream evaluation metrics. We can access them via the *@* operator 
```{r accessors, eval=TRUE}
dim(DATA@refCounts)
head(DATA@refcolData)
```

## Select domain-informative genes

`featureSelection()` identifies genes informative for each spatial domain. These
genes are used to estimate and simulate domain-specific expression patterns.

```{r feature-selection, eval = TRUE}
DATA <- featureSelection(
  spaDesign = DATA,
  logfc_cutoff = 0.7,
  mean_in_cutoff = 1.8,
  max_num_gene = 10,
  n_cores = 4
)

lapply(topGenes(DATA), head)
```

The thresholds could be tuned. A stricter `logfc_cutoff`
selects genes with stronger domain contrast, while a higher `mean_in_cutoff`
filters genes that are weakly expressed inside the domain.

## Estimate model parameters

`spaDesign` estimates two complementary parts of the spatial transcriptomics
signal:

- `estimation_FGEM()` learns domain location spatial distribution patterns.
- `estimation_NNGP()` learns spatial expression patterns for selected genes.

```{r parameter-estimation}
DATA <- estimation_FGEM(
  spaDesign = DATA,
  iter_max = 1000,
  M_candidates = 2:6,
  tol = 0.1,
  n_cores = 4,
  verbose = FALSE
)

DATA <- estimation_NNGP(
  spaDesign = DATA,
  n_neighbors = 10,
  order = "AMMD",
  X = NULL,
  verbose = FALSE
)
```

After this step, the object contains the fitted parameters needed for
simulation.


The fitted object contains selected genes and model parameters.

```{r load-fitted-example, include=FALSE, eval=TRUE}
DATA <- readRDS(system.file("extdata", "ref_humanBrain.rds", package = "spaDesign"))
```

Specifically, the estimated NNGP parameters are stored by domain and gene. For example,
`Layer1` and gene `FTH1` contain an estimated regression coefficient:

```{r show-nngp-output, echo=FALSE, eval=TRUE}
DATA@paramsGP$Layer1$FTH1$Beta
```

The estimated FGEM parameters are stored by domain. For example, the mixing
proportions from the AIC-selected model for `Layer1` are:

```{r show-fgem-output, echo=FALSE, eval=TRUE}
DATA@paramsFG$Layer1$best_model_AIC$pi
```

## Power analysis over sequencing depth and effect size

`powerAnalysisEffectSize()` simulates datasets over a grid of sequencing depth
scaling factors and effect size scaling factors. Domain recovery is evaluated by
NMI after clustering the simulated data.

```{r effect-size-power}
effect_results <- powerAnalysisEffectSize(
  spaDesign = DATA,
  es_range = c(0.5, 1, 2),
  seq_depth_range = c(0.5, 1, 2, 3, 5, 7, 10),
  n_rep = 3,
  n_cores = 4
)

head(effect_results)
```


```{r effect-size-power-values, eval=TRUE, include=FALSE}
effect_results <- data.frame(
  seq_depth = c(0.5, 1, 2, 3, 5, 7),
  effect_size = c(1, 1, 1, 1, 1, 1),
  total_counts = c(640829, 1282539, 2566330, 3848292, 6411184, 8973274),
  NMI = c(0.42, 0.61, 0.74, 0.80, 0.84, 0.85)
)
```

```{r effect-size-power-output, echo=FALSE, eval=TRUE}
head(effect_results)
```
The output contains one row per simulation replicate and scenario:

- `seq_depth`: relative sequencing depth compared with the pilot dataset
- `effect_size`: signal scaling factor
- `total_counts`: total counts in the simulated dataset
- `NMI`: agreement between inferred and true domain labels

If the pilot experiment has an observed depth in millions of reads or UMIs, the
relative scale can be converted to an absolute scale:

```{r absolute-depth}
pilot_depth <- sum(refCounts(DATA)) / 1e6
effect_results$absolute_depth <- effect_results$seq_depth * pilot_depth
```

## Power analysis under spatial perturbation

Spatial perturbation analysis evaluates robustness when a fraction of genes
deviate from the undisturbed spatial pattern.

```{r spatial-power}
spatial_results <- powerAnalysisSpatial(
  spaDesign = DATA,
  SIGMA = 1,
  prop_range = c(0.6, 0.8, 1.0),
  seq_depth_range = c(0.5, 1, 2, 3, 5, 7, 10),
  n_rep = 5,
  n_cores = 4
)

head(spatial_results)
```

Here, `prop_range` controls the proportion of genes that retain the original
spatial pattern, and `SIGMA` controls the spatial noise level.

## Detect sequencing depth saturation

The saturation point is estimated by fitting a monotone smooth curve between
sequencing depth and NMI. The detected saturation depth is the first point where
the fitted curve has sufficiently small marginal improvement after reaching a
specified fraction of its maximum performance.

```{r saturation-effect,eval = T}
sat_effect <- saturationDetection(
  data = effect_results,
  metric_col = "NMI",
  depth_col = "seq_depth",
  group_cols = "effect_size",
  pilot_depth = NULL,
  slope_threshold = 0.05,
  required_metric_percentage = 0.8,
  aggregate_reps = TRUE
)

sat_effect$summary
```

Important tuning parameters:

- `slope_threshold`: lower values require a flatter curve before declaring
  saturation.
- `required_metric_percentage`: higher values require the curve to get closer
  to its maximum before saturation is considered.
- `aggregate_reps`: when `TRUE`, replicate results at the same sequencing depth
  are averaged before curve fitting.

Visualize the fitted curves and detected saturation depths:

```{r plot-saturation, eval = T}
plotSaturation(
  sat_effect,
  metric_col = "NMI",
  use_absolute = TRUE,
  facet = TRUE,
  label_saturation = TRUE
)
```

## Interactive Shiny workflow

The package includes a Shiny application with two main steps:

1. Data preparation: choose a built-in reference dataset or upload SpaceRanger
   output files.
2. Sequencing depth estimation: run baseline, effect-size, or spatial
   perturbation simulations and visualize the saturation curve.

Launch the app from an installed package:

```{r shiny-app}
library(spaDesign)

runSpaDesignApp()
```

For uploaded data, the app expects:

- `filtered_feature_bc_matrix.h5`
- `tissue_positions_list.csv`
- optional annotation CSV with `barcode` and `domain` columns

If no annotation file is provided, the app estimates spatial domains from the
uploaded count matrix.

## A minimal end-to-end template

The following template summarizes the full analysis:

```{r full-template}
library(spaDesign)

example_data <- readRDS(system.file("extdata", "exampleData.rds", package = "spaDesign"))

DATA <- createDesignObject(
  count_matrix = example_data$count,
  loc = example_data$loc
)

DATA <- featureSelection(
  DATA,
  logfc_cutoff = 0.7,
  mean_in_cutoff = 1.8,
  max_num_gene = 10,
  n_cores = 4
)

DATA <- estimation_FGEM(
  DATA,
  iter_max = 1000,
  M_candidates = 2:6,
  tol = 0.1,
  n_cores = 4,
  verbose = FALSE
)

DATA <- estimation_NNGP(
  DATA,
  n_neighbors = 10,
  order = "AMMD",
  X = NULL,
  verbose = FALSE
)

rst <- powerAnalysisEffectSize(
  DATA,
  es_range = c(1, 2),
  seq_depth_range = c(0.5, 1, 2, 3, 5, 7, 10),
  n_rep = 5,
  n_cores = 4
)

pilot_depth <- sum(refCounts(DATA)) / 1e6

sat <- saturationDetection(
  rst,
  group_cols = "effect_size",
  pilot_depth = pilot_depth,
  aggregate_reps = TRUE
)

sat$summary
plotSaturation(sat)
```

## Practical recommendations

Use a pilot dataset that is representative of the tissue, chemistry, and
expected domain structure of the planned study. The saturation depth estimated
by `spaDesign` is conditional on this pilot signal.

Start with a modest grid such as `seq_depth_range = c(0.5, 1, 2, 3, 5, 7, 10)`
and increase `n_rep` after confirming that the pipeline runs successfully.

When comparing multiple experimental conditions, interpret the recommended
depth alongside the achieved NMI. A low saturation depth with a low maximum NMI
may indicate that sequencing depth is not the main limiting factor.
